Modeling browser-based distributed evolutionary computation systems

### 2015-02-08 modeling-volunteer-computing-gecco.tex
Commented results and modified a reference
Using a book chapter instead of a arxiv pre-print. Also, commented some parts in the results and conclusions sections.

The server was also written using node.js and the {\tt express.js}
module and has three REST \cite{DBLP:journals/corr/abs-1105-4971}
functions, that is, three functions that can be accessed from any
client, including the browser, via a request encoded in an URL and a
standard HTTP command. These functions are\begin{itemize}
\item {\tt GET random} returns a random, non-evaluated individual,
  from the pool. The pool is just an array containing chromosomes,
  that is initialized at the beginning of the experiment.
\item {\tt PUT one} puts a single individual in the pool. In REST
  conventions, {\tt PUT} is used to create a resource, that is why it
  is used, in this case, to add a new individual to the pool.
\item {\tt GET log} returns the experiment log so far. It is used by
  the researcher to gather data, and as a transparency measure so that
  anyone can use it. 
\end{itemize}

@@

The litmus test that all experiments should pass, that is, that they
work and found the solution, passed. In fact, the first version of the
algorithm (until January 2015) used too much memory on the browser,
which crashed after a while running. A total of 7 experiments were
done with this version. After January, a new version of the
evolutionary algorithm with non-cached fitness was used which did not
crash the browser and was thus allowed to run as much time as
needed. That was the version used in a lightning talk at the beginning
of February. A total of 10 experiments were made with this second
version. In the first version, finding the solution took a few
minutes. The second needed around 20 minutes. In the logs used, it was
not checked that the algorithm had found the solution, but in most
cases it had. More systematic experimentation would have to wait until
the next version, since in this first one we were looking for some
kind of systematicity in the behavior of users and the experimental
apparatus is the minimum possible to have it running.
%
\begin{figure*}[htb]
        \centering
        \subfigure[First experiment]{
                \includegraphics[width=0.45\linewidth]{data/diffs-ips-1.png}
                \label{fig:e1_p}
        }
        \subfigure[Second experiment]{
                \includegraphics[width=0.45\linewidth]{data/diffs-ips-2.png}
                \label{fig:e2_p}
        }
        \caption{Histogram of number of different IPs in an experiment}\label{fig:histo:ips}
\end{figure*}

The fact that it run correctly during a massive lightning talk means that this setup is a
valid, no-cost, platform for massive evolutionary algorithms. There
were several ways of introducing it: during a talk, telling the
audience to visit the web that was posted in the first slide, and
after the talk, inserting the URL in a tweet with a brief explanation
and several tweets more indicating what it was about and what it
did. If any question arose, it was answered as fast as possible. In
the first experiment, the median number of different IPs was 18, with
a maximum of 28 and a minimum of 6. In the second, the median was 12.5
with a maximum of 19 and minimum of 8. The histogram of different IPs
in each experiment is shown in figure \ref{fig:histo:ips}. Probably
the most remarkable thing is that a cluster of 6 computers can, as a
minimum, be gathered for a distributed computing experiment by just
announcing it as a tweet, but the fact that we can obtain more than a
dozen computers in more than half the cases is also remarkable, and
goes to prove the excess of computing resources that people,
willingly, lend to a simple experiment.
%
\begin{figure*}[htb]
        \centering
        \subfigure[First experiment, run 3]{
                \includegraphics[width=0.45\linewidth]{data/log-puts-1-exp7.png}
                \label{fig:d1}
        }
        \subfigure[Second experiment, run 7]{
                \includegraphics[width=0.45\linewidth]{data/log-puts-2-traps-eps-3.png}
                \label{fig:d2}
        }
        \caption{IPs and the number of {\tt PUT}s (one every $n$
          generations, 50 in the first experiment, 100 in the second
          experiment) they did, sorted. These correspond to the runs
          in each batch of experiments with the max number of
          different IPs. Please note the algorithmic scale in the $x$ axis.}\label{fig:dist}
\end{figure*}

@@

At any rate, this exponential trend in the first posts breaks down
after a number of IPs. That could be due to several possible factors:
persons checking the web for just an instant, late comers, or very
slow computers that are not able to lend more cycles to the
simulation. The first IP in both cases contributes on the order of a
hundred PUTs, that is, several thousands generations and several
hundred thousands evaluations, which, once again, is quite remarkable
for such a simple experiment announced by an ephemeral tweet. It
should be noted that, since the speed in the browser virtual machines
(Chrome, mainly, but also Firefox) is as fast as the node.js version,
a parallel version does not add overhead to the single-browser
version, which, as mentioned above, can run in an independent
way: a second browser always adds to the first, even if the total
number of evaluations and the time needed will not speed up in the
same way. However, even taking into account the stochastic nature of a
single evolutionary algorithm, what we are going to find is a high
dispersion of the times needed to reach a solution, which are plotted
in Figure \ref{fig:t} against the number of IPs. There is a wide
range of durations, although there is no clear decrease of the time
needed as the number of IPs used increase. However, we will have to
investigate further the speed ups achieved and if these are
algorithmic or due to the particular asynchronous and web-based
implementation. Figure \ref{fig:intervals} might help understand the
kind of raw material we are using: It plots an histogram of the time
needed by all experiments to process 100 generations, 25600
evaluations and the associated evolutionary geneticry. The vast
majority of browsers processes it in less than 3 seconds (logarithm(
time ) < 0.5, which is equivalent to 3.16). However, it is interesting
to note that there are some intervals that go up to the thousands of
seconds, for which we cannot really offer an explanation, other than
the user leaving the page and returning after a while. However, 75\%
of the cases take less than 4 seconds, which is thus a statistical
measure of the kind of performance we can expect from clients for this
type of problem. We can also conclude that 75\%, or a majority, will
be in the fastest tier, although there will be a small amount of them
that will be quite slow and will probably be impossible to accomodate
in the distributed computing experiment.

@@

>>>>

The server was also written using node.js and the {\tt express.js}
module and has three REST \cite{Castillo12REST} %FERGU: cito a este paper que es un capítulo de libro en vez de Arxiv
functions, that is, three functions that can be accessed from any
client, including the browser, via a request encoded in an URL and a
standard HTTP command. These functions are\begin{itemize}
\item {\tt GET random} returns a random, non-evaluated individual,
  from the pool. The pool is just an array containing chromosomes,
  that is initialized at the beginning of the experiment.
\item {\tt PUT one} puts a single individual in the pool. In REST
  conventions, {\tt PUT} is used to create a resource, that is why it
  is used, in this case, to add a new individual to the pool.
\item {\tt GET log} returns the experiment log so far. It is used by
  the researcher to gather data, and as a transparency measure so that
  anyone can use it. 
\end{itemize}

@@

The litmus test that all experiments should pass, that is, that they
work and found the solution, passed. %FERGU: que quiere decir esta primera frase?
In fact, the first version of the
algorithm (until January 2015) used too much memory on the browser,
which crashed after a while running. A total of 7 experiments were
done with this version. After January, a new version of the
evolutionary algorithm with non-cached fitness was used which did not
crash the browser and was thus allowed to run as much time as
needed. That was the version used in a lightning talk at the beginning
of February. A total of 10 experiments were made with this second
version. In the first version, finding the solution took a few
minutes. The second needed around 20 minutes. In the logs used, it was
not checked that the algorithm had found the solution, but in most
cases it had. More systematic experimentation would have to wait until
the next version, since in this first one we were looking for some
kind of systematicity in the behavior of users and the experimental
%FERGU: systematicity? systematization mejor?
apparatus is the minimum possible to have it running.
%
\begin{figure*}[htb]
        \centering
        \subfigure[First experiment]{
                \includegraphics[width=0.45\linewidth]{data/diffs-ips-1.png}
                \label{fig:e1_p}
        }
        \subfigure[Second experiment]{
                \includegraphics[width=0.45\linewidth]{data/diffs-ips-2.png}
                \label{fig:e2_p}
        }
        \caption{Histogram of number of different IPs in an experiment}\label{fig:histo:ips}
\end{figure*}

The fact that it run correctly during a massive lightning talk means that this setup is a
valid, no-cost, platform for massive evolutionary algorithms. There
were several ways of introducing it: during a talk, telling the
audience to visit the web that was posted in the first slide, and
after the talk, inserting the URL in a tweet with a brief explanation
and several tweets more indicating what it was about and what it
did. If any question arose, it was answered as fast as possible. In
the first experiment, the median number of different IPs was 18, with
a maximum of 28 and a minimum of 6. In the second, the median was 12.5
with a maximum of 19 and minimum of 8. The histogram of different IPs
in each experiment is shown in figure \ref{fig:histo:ips}. Probably
the most remarkable thing is that a cluster of 6 computers can, as a
minimum, be gathered for a distributed computing experiment by just
announcing it as a tweet, but the fact that we can obtain more than a
dozen computers in more than half the cases is also remarkable, and
goes to prove the excess of computing resources that people,
willingly, lend to a simple experiment.
%FERGU: dí que la persona que lo anunció en twitter tiene X seguidores
%
\begin{figure*}[htb]
        \centering
        \subfigure[First experiment, run 3]{
                \includegraphics[width=0.45\linewidth]{data/log-puts-1-exp7.png}
                \label{fig:d1}
        }
        \subfigure[Second experiment, run 7]{
                \includegraphics[width=0.45\linewidth]{data/log-puts-2-traps-eps-3.png}
                \label{fig:d2}
        }
        \caption{IPs and the number of {\tt PUT}s (one every $n$
          generations, 50 in the first experiment, 100 in the second
          experiment) they did, sorted. These correspond to the runs
          in each batch of experiments with the max number of
          different IPs. Please note the algorithmic scale in the $x$ axis.}\label{fig:dist}
\end{figure*}

@@

At any rate, this exponential trend in the first posts breaks down
after a number of IPs. That could be due to several possible factors:
persons checking the web for just an instant, late comers, or very
slow computers that are not able to lend more cycles to the
simulation. The first IP in both cases contributes on the order of a
hundred PUTs, that is, several thousands generations and several
hundred thousands evaluations, which, once again, is quite remarkable
for such a simple experiment announced by an ephemeral tweet. It
should be noted that, since the speed in the browser virtual machines
(Chrome, mainly, but also Firefox) is as fast as the node.js version,
a parallel version does not add overhead to the single-browser
version, which, as mentioned above, can run in an independent
way: a second browser always adds to the first, even if the total
number of evaluations and the time needed will not speed up in the
same way. However, even taking into account the stochastic nature of a
single evolutionary algorithm, what we are going to find is a high
dispersion of the times needed to reach a solution, which are plotted
in Figure \ref{fig:t} against the number of IPs. There is a wide
range of durations, although there is no clear decrease of the time
needed as the number of IPs used increase. However, we will have to
investigate further the speedups achieved and if these are %FERGU: speedup va junto
algorithmic or due to the particular asynchronous and web-based
implementation. Figure \ref{fig:intervals} might help understand the
kind of raw material we are using: It plots an histogram of the time
needed by all experiments to process 100 generations, 25600
evaluations and the associated evolutionary geneticry. The vast %FERGU: geneticry?
majority of browsers processes it in less than 3 seconds (logarithm( 
time ) < 0.5, which is equivalent to 3.16). However, it is interesting
to note that there are some intervals that go up to the thousands of
seconds, for which we cannot really offer an explanation, other than
the user leaving the page and returning after a while. However, 75\%
of the cases take less than 4 seconds, which is thus a statistical
measure of the kind of performance we can expect from clients for this
type of problem. We can also conclude that 75\%, or a majority, will
be in the fastest tier, although there will be a small amount of them
that will be quite slow and will probably be impossible to accommodate
in the distributed computing experiment.

@@

%FERGU: Creo que te falta terminar las conclusiones, pero por si acaso habría que decir: Este tipo de computación se basa en tal y tal y tiene tales problemas. Nuestra metodología ha propuesto usar tal tecnología por tal y tal. Los resultados demuestran que los usuarios hacen tal y tal y que por lo tanto hay que tener en cuenta el tal. Bueno, qué te voy a contar...
